---
title: "MONECA Algorithm Comparison: moneca() vs moneca_fast()"
author: "Jonas Touboel and Anton Grau Larsen"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_width: 8
    fig_height: 6
vignette: >
  %\VignetteIndexEntry{MONECA Algorithm Comparison: moneca() vs moneca_fast()}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "figures/",
  warning = FALSE,
  message = FALSE
)

# Check if required packages are available
required_packages <- c("ggplot2", "ggraph", "igraph", "dplyr", "tidygraph", "Matrix")
missing_packages <- required_packages[!sapply(required_packages, requireNamespace, quietly = TRUE)]

if (length(missing_packages) > 0) {
  knitr::opts_chunk$set(eval = FALSE)
  message("Some required packages are missing: ", paste(missing_packages, collapse = ", "))
  message("Code examples will not be evaluated.")
}
```

## Introduction

The moneca package provides two main algorithms for mobility network clustering analysis: the original `moneca()` function and the optimized `moneca_fast()` function. This vignette provides a detailed comparison of both algorithms, explaining their internal workflows, differences, and trade-offs to help users choose the most appropriate method for their analysis.

Both algorithms implement the same core MONECA methodology but with different optimization strategies and computational approaches. Understanding these differences is crucial for selecting the right tool based on your data size, computational resources, and accuracy requirements.

## Overview of the MONECA Methodology

Before diving into the specific implementations, it's important to understand the core MONECA methodology that both algorithms implement:

1. **Relative Risk Calculation**: Convert mobility tables to relative risk matrices
2. **Network Construction**: Create weighted networks from relative risk values
3. **Clique Detection**: Identify complete subgraphs (cliques) in the network
4. **Segmentation**: Group nodes into segments based on clique membership
5. **Aggregation**: Create new mobility tables aggregated by segments
6. **Iteration**: Repeat the process for multiple hierarchical levels

The key differences between `moneca()` and `moneca_fast()` lie in how they implement steps 3 (clique detection) and 4 (segmentation).

## The Original moneca() Algorithm

### Workflow Overview

The original `moneca()` function implements a conservative, thorough approach that prioritizes accuracy and reproducibility. Here's the detailed workflow:

```{r eval=FALSE}
# Load the package and create sample data
library(moneca)
mobility_data <- generate_mobility_data(n_classes = 6, seed = 123)

# Run the original algorithm
result_original <- moneca(mobility_data, segment.levels = 3)
```

### Step-by-Step Workflow

#### Step 1: Input Validation and Preprocessing
```{r eval=FALSE}
# The function begins with input validation
mx <- mobility_data  # Input mobility table
segment.levels <- 3  # Number of hierarchical levels
cut.off <- 1         # Minimum relative risk threshold
```

The algorithm expects a square mobility table with row and column totals in the last row and column.

#### Step 2: Relative Risk Matrix Calculation
```{r eval=FALSE}
# Internal function: weight.matrix()
# Calculates observed/expected ratios for each cell
l <- nrow(mx)
o.r.s <- mx[-l, l]        # Origin totals
o.c.s <- mx[l, -l]        # Destination totals
total.total <- mx[l,l]    # Grand total

# Expected values under independence
row.share <- o.r.s/total.total
col.share <- o.c.s/total.total
total.mobility <- sum(mx[-l,-l])
mx.1_exp <- row.share %*% t(col.share) * total.mobility

# Relative risks
mx.1_net <- mx[-l,-l] / mx.1_exp
mx.1i <- as.matrix(mx.1_net)
mx.1i[mx.1i < cut.off] <- NA  # Apply threshold
mx.1i <- mx.1i + t(mx.1i)     # Make symmetric
diag(mx.1i) <- NA             # Remove diagonal
```

#### Step 3: Network Construction
```{r eval=FALSE}
# Convert to adjacency matrix for graph creation
mx.1i.graph <- mx.1i
mx.1i.graph[is.na(mx.1i.graph)] <- 0

# Create undirected, weighted graph
gra.1ii <- moneca_graph_from_adjacency(
  adjmatrix = mx.1i.graph, 
  mode = "undirected", 
  weighted = TRUE, 
  diag = FALSE
)
```

#### Step 4: Clique Detection
```{r eval=FALSE}
# Find ALL cliques in the network
clique <- moneca_cliques(gra.1ii)
```

**Key characteristic**: The original algorithm finds ALL cliques in the network, including overlapping ones. This provides comprehensive coverage but can be computationally expensive for dense networks.

#### Step 5: Edge Processing and Segmentation
```{r eval=FALSE}
# The find.segments() function processes edges in order of strength
# Extract edges and sort by weight (strongest first)
edges <- which(!is.na(mat) & mat > cut.off, arr.ind = TRUE)
edge_weights <- mat[edges]
edge_order <- order(edge_weights, decreasing = TRUE)
edges <- edges[edge_order, , drop = FALSE]

# Initialize group assignments
group <- vector(mode = "numeric", length = nrow(mat))
names(group) <- rownames(mat)

# Process each edge in order of decreasing strength
for (i in seq_len(nrow(edges))) {
  edge <- edges[i, ]
  current_groups <- group[edge]
  
  # Determine group assignment based on existing memberships
  # and clique constraints
  if (sum(current_groups) == 0) {
    # Both nodes unassigned: create new group
    group[edge] <- i  
  } else {
    # At least one node assigned: test clique membership
    group.candidates <- current_groups[current_groups != 0] 
    group.members <- which(group %in% group.candidates)
    potential.clique <- unique(sort(c(group.members, edge)))
    
    # Test if potential group forms a valid clique
    test <- clique.test(cliques, potential.clique)
    if (test == TRUE) {
      group[potential.clique] <- group.assigned             
    }
  }
}
```

**Key characteristics**:
- **Sequential processing**: Edges are processed one by one in order of strength
- **Clique validation**: Every potential group is tested against all known cliques
- **Conservative grouping**: Groups are only formed if they satisfy clique constraints

#### Step 6: Matrix Aggregation
```{r eval=FALSE}
# Aggregate the original mobility table by the identified segments
groups.1 <- c(segments$membership, length(segments$membership) + 1)
mx.2_r <- rowsum(mx, groups.1)
mx.2_r_t <- t(mx.2_r)
mx.2_rc_t <- rowsum(mx.2_r_t, groups.1)
mx.2g <- t(mx.2_rc_t)
```

#### Step 7: Hierarchical Iteration
```{r eval=FALSE}
# Repeat the process for the aggregated matrix
# Continue until segment.levels is reached or no further segmentation possible
for (i in 2:segment.levels) {
  segments <- make.segments(mx.2g, cut.off, mode, delete.upper.tri, small.cell.reduction)
  mx.2g <- segment.matrix(mx.2g, segments)
  # ... store results for level i
}
```

### Advantages of moneca()

1. **Accuracy**: Finds all cliques, ensuring no potential segments are missed
2. **Deterministic**: Results are completely reproducible
3. **Thorough**: Conservative approach minimizes false segmentation
4. **Well-tested**: Extensively validated algorithm with proven track record
5. **Comprehensive**: Handles edge cases and complex network structures reliably

### Disadvantages of moneca()

1. **Computational complexity**: O(3^n) in worst case for clique enumeration
2. **Memory intensive**: Stores all cliques in memory
3. **Slow for large networks**: Performance degrades significantly with network size
4. **No early stopping**: Processes all edges even when further segmentation is unlikely

## The Optimized moneca_fast() Algorithm

### Workflow Overview

The `moneca_fast()` function implements several optimization strategies while maintaining the core MONECA methodology. It prioritizes computational efficiency while preserving algorithmic accuracy for most practical use cases.

```{r eval=FALSE}
# Run the optimized algorithm
result_fast <- moneca_fast(
  mobility_data, 
  segment.levels = 3,
  use.sparse = FALSE,        # Optional sparse matrix support
  min.density = 0.01,        # Early stopping threshold
  max.clique.size = NULL,    # Clique size limitation
  progress = TRUE            # Progress monitoring
)
```

### Step-by-Step Workflow

#### Step 1: Enhanced Input Validation and Configuration
```{r eval=FALSE}
# Additional parameters for optimization
use.sparse <- FALSE           # Sparse matrix support
min.density <- 0.01          # Minimum edge density for continuation
max.clique.size <- NULL      # Maximum clique size to consider
progress <- TRUE             # Progress bar display
```

#### Step 2: Optimized Relative Risk Calculation
```{r eval=FALSE}
# Fast weight matrix calculation with vectorized operations
weight.matrix.fast <- function(mx, cut.off = 1, symmetric = TRUE, 
                               diagonal = NULL, small.cell.reduction = 0) {
  l <- nrow(mx)
  
  # Vectorized operations for better performance
  o.r.s <- mx[-l, l]
  o.c.s <- mx[l, -l]
  total.total <- mx[l, l]
  row.share <- o.r.s / total.total
  col.share <- o.c.s / total.total
  total.mobility <- sum(mx[-l, -l])
  
  # Vectorized expected calculation using outer product
  mx.1_exp <- outer(row.share, col.share) * total.mobility
  
  # Continue with relative risk calculation...
}
```

**Optimization**: Uses vectorized operations and `outer()` for more efficient matrix calculations.

#### Step 3: Sparse Matrix Support (Optional)
```{r eval=FALSE}
# Convert to sparse matrix if requested and beneficial
if (use.sparse && requireNamespace("Matrix", quietly = TRUE)) {
  if (!inherits(mx, "sparseMatrix")) {
    mx <- Matrix::Matrix(mx, sparse = TRUE)
  }
}
```

**Optimization**: Optional sparse matrix support for large, sparse datasets.

#### Step 4: Maximal Clique Detection
```{r eval=FALSE}
# Use MAXIMAL cliques instead of ALL cliques
if (is.null(max.clique.size)) {
  cliques <- moneca_max_cliques(graph, min = 2)
} else {
  cliques <- moneca_max_cliques(graph, min = 2, max = max.clique.size)
}
```

**Key optimization**: Uses maximal cliques instead of all cliques. Maximal cliques are cliques that cannot be extended by adding another vertex, which significantly reduces the number of cliques to consider while maintaining algorithmic soundness.

#### Step 5: Early Stopping Based on Density
```{r eval=FALSE}
# Early stopping if graph becomes too sparse
density <- igraph::edge_density(graph)
if (density < min.density) {
  n <- nrow(mat)
  out <- list(
    membership = as.factor(1:n),
    cliques = as.list(1:n)
  )
  return(out)  # Return trivial segmentation
}
```

**Optimization**: Stops processing when network density falls below threshold, indicating limited potential for meaningful segmentation.

#### Step 6: Vectorized Edge Processing
```{r eval=FALSE}
# Vectorized edge extraction and sorting
edges <- which(!is.na(mat) & mat > cut.off, arr.ind = TRUE)
edge_weights <- mat[edges]
edge_order <- order(edge_weights, decreasing = TRUE)
edges <- edges[edge_order, , drop = FALSE]

# Pre-compute clique membership matrix for fast testing
clique_membership <- matrix(FALSE, nrow = length(cliques), ncol = nrow(mat))
for (i in seq_along(cliques)) {
  clique_membership[i, cliques[[i]]] <- TRUE
}
```

**Optimization**: Pre-computes clique membership matrix for O(1) clique testing instead of repeated searches.

#### Step 7: Optimized Segmentation Algorithm
```{r eval=FALSE}
# Fast clique test using pre-computed membership
is_clique <- any(apply(clique_membership[, potential_clique, drop = FALSE], 1, all))

if (is_clique) {
  # Merge groups - use the smallest group number
  target_group <- min(assigned_groups)
  group[potential_clique] <- target_group
}
```

**Optimization**: Uses matrix operations for clique testing instead of iterative searches.

#### Step 8: Optimized Level Processing
```{r eval=FALSE}
# Optimized level.down function with vectorized operations
level.down.fast <- function(level.current, level.below) {
  # Remove isolates efficiently
  lengths <- lengths(level.current)
  level.current <- level.current[lengths > 1]
  
  if (length(level.current) == 0) return(list())
  
  # Vectorized operation using lapply
  ud <- lapply(level.current, function(d) unlist(level.below[d]))
  return(ud)
}
```

**Optimization**: Uses vectorized operations and `lengths()` for more efficient processing.

### Advantages of moneca_fast()

1. **Performance**: Significantly faster, especially for larger networks
2. **Memory efficient**: Uses maximal cliques to reduce memory footprint
3. **Early stopping**: Avoids unnecessary computation when segmentation potential is low
4. **Sparse matrix support**: Handles large sparse matrices efficiently
5. **Progress monitoring**: Provides feedback for long-running analyses
6. **Scalable**: Better performance scaling with network size

### Disadvantages of moneca_fast()

1. **Potentially less comprehensive**: Maximal cliques might miss some valid segments
2. **Early stopping trade-offs**: May terminate segmentation prematurely in some cases
3. **Complexity**: More parameters to tune (density thresholds, clique size limits)
4. **Less tested**: Newer algorithm with less extensive validation
5. **Parameter sensitivity**: Results may be more sensitive to parameter choices

## Detailed Algorithm Comparison

### Clique Detection Strategies

**moneca()**: Uses ALL cliques
```{r eval=FALSE}
# Finds every possible clique in the network
cliques <- moneca_cliques(graph)
# For a 6-node complete graph: finds 2^6 - 6 - 1 = 57 cliques
```

**moneca_fast()**: Uses MAXIMAL cliques
```{r eval=FALSE}
# Finds only maximal cliques (cannot be extended)
cliques <- moneca_max_cliques(graph, min = 2)
# For a 6-node complete graph: finds only 1 maximal clique
```

### Computational Complexity

| Aspect | moneca() | moneca_fast() |
|--------|----------|---------------|
| Clique enumeration | O(3^n) worst case | O(d^3) where d is degeneracy |
| Memory usage | O(2^n) for storing cliques | O(k) where k is number of maximal cliques |
| Edge processing | O(m × c) where c is clique count | O(m × log(c)) with pre-computed lookup |
| Overall complexity | Exponential in worst case | Polynomial in most practical cases |

### Performance Characteristics

```{r eval=FALSE}
# Benchmarking example (not run in vignette)
library(microbenchmark)

# Small network (6 nodes)
small_data <- generate_mobility_data(n_classes = 6, seed = 123)
microbenchmark(
  original = moneca(small_data, segment.levels = 2),
  fast = moneca_fast(small_data, segment.levels = 2, progress = FALSE),
  times = 10
)

# Medium network (12 nodes)
medium_data <- generate_mobility_data(n_classes = 12, seed = 123)
microbenchmark(
  original = moneca(medium_data, segment.levels = 2),
  fast = moneca_fast(medium_data, segment.levels = 2, progress = FALSE),
  times = 5
)
```

Expected performance differences:
- **Small networks (< 10 nodes)**: Minimal difference, sometimes original is faster due to lower overhead
- **Medium networks (10-20 nodes)**: Fast version typically 2-5x faster
- **Large networks (> 20 nodes)**: Fast version can be 10-100x faster or more

## Practical Examples

### Example 1: Small Dataset Comparison

```{r small-example}
library(moneca)

# Generate a small mobility dataset
small_data <- generate_mobility_data(n_classes = 5, immobility_strength = 0.8, seed = 42)
print(small_data)

# Run both algorithms
result_original <- moneca(small_data, segment.levels = 2)
result_fast <- moneca_fast(small_data, segment.levels = 2, progress = FALSE)

# Compare results
cat("Original algorithm segments at level 2:\n")
print(result_original$segment.list[[2]])

cat("\nFast algorithm segments at level 2:\n")
print(result_fast$segment.list[[2]])

# Check if results are identical
identical_results <- identical(result_original$segment.list, result_fast$segment.list)
cat("\nResults identical:", identical_results, "\n")
```

### Example 2: Performance Comparison

```{r performance-example}
# Generate medium-sized dataset
medium_data <- generate_mobility_data(n_classes = 8, seed = 123)

# Time both algorithms
cat("Timing original algorithm...\n")
time_original <- system.time({
  result_orig <- moneca(medium_data, segment.levels = 2)
})

cat("Timing fast algorithm...\n")
time_fast <- system.time({
  result_fast <- moneca_fast(medium_data, segment.levels = 2, progress = FALSE)
})

cat("Original algorithm time:", time_original["elapsed"], "seconds\n")
cat("Fast algorithm time:", time_fast["elapsed"], "seconds\n")
cat("Speedup factor:", round(time_original["elapsed"] / time_fast["elapsed"], 2), "x\n")
```

### Example 3: Different Segmentation Outcomes

Sometimes the algorithms may produce different results due to their different clique detection strategies:

```{r different-outcomes}
# Create a dataset where algorithms might differ
complex_data <- generate_mobility_data(
  n_classes = 6, 
  immobility_strength = 0.6,  # Lower immobility for more complex structure
  class_clustering = 0.3,      # Some clustering
  noise_level = 0.1,          # Some noise
  seed = 789
)

# Run both algorithms
seg_original <- moneca(complex_data, segment.levels = 3)
seg_fast <- moneca_fast(complex_data, segment.levels = 3, progress = FALSE)

# Compare segment structures
cat("Original algorithm - Level 2 segments:\n")
lapply(seg_original$segment.list[[2]], function(x) paste(x, collapse = ", "))

cat("\nFast algorithm - Level 2 segments:\n") 
lapply(seg_fast$segment.list[[2]], function(x) paste(x, collapse = ", "))

# Compare number of segments at each level
levels_orig <- sapply(seg_original$segment.list, length)
levels_fast <- sapply(seg_fast$segment.list, length)

comparison_df <- data.frame(
  Level = 1:length(levels_orig),
  Original = levels_orig,
  Fast = levels_fast,
  Difference = levels_orig - levels_fast
)
print(comparison_df)
```

## Parameter Tuning for moneca_fast()

The `moneca_fast()` function provides several parameters for performance tuning:

### min.density Parameter

Controls early stopping based on network density:

```{r density-tuning}
# Test different density thresholds
data_test <- generate_mobility_data(n_classes = 6, seed = 456)

# Conservative (lower threshold - more processing)
result_conservative <- moneca_fast(data_test, min.density = 0.001, progress = FALSE)

# Aggressive (higher threshold - less processing)  
result_aggressive <- moneca_fast(data_test, min.density = 0.05, progress = FALSE)

cat("Conservative approach - levels:", length(result_conservative$segment.list), "\n")
cat("Aggressive approach - levels:", length(result_aggressive$segment.list), "\n")
```

### max.clique.size Parameter

Limits maximum clique size to control computation:

```{r clique-size-tuning}
# Unlimited clique size
result_unlimited <- moneca_fast(data_test, max.clique.size = NULL, progress = FALSE)

# Limited clique size
result_limited <- moneca_fast(data_test, max.clique.size = 4, progress = FALSE)

cat("Unlimited cliques - segments at level 2:", length(result_unlimited$segment.list[[2]]), "\n")
cat("Limited cliques - segments at level 2:", length(result_limited$segment.list[[2]]), "\n")
```

### Sparse Matrix Support

For large, sparse matrices:

```{r sparse-support}
# Note: This example demonstrates the parameter usage
# In practice, sparse matrices are beneficial for very large datasets

# Regular matrix processing
result_dense <- moneca_fast(data_test, use.sparse = FALSE, progress = FALSE)

# Sparse matrix processing (if Matrix package available)
if (requireNamespace("Matrix", quietly = TRUE)) {
  result_sparse <- moneca_fast(data_test, use.sparse = TRUE, progress = FALSE)
  cat("Sparse matrix support available and tested\n")
} else {
  cat("Matrix package not available - sparse support skipped\n")
}
```

## Decision Guidelines

### When to Use moneca()

Choose the original `moneca()` function when:

1. **Accuracy is paramount**: You need the most comprehensive and conservative segmentation
2. **Small to medium datasets**: Network size is manageable (< 15 nodes typically)
3. **Reproducibility is critical**: You need identical results across different runs and platforms
4. **Methodological consistency**: You're replicating or extending previous research using MONECA
5. **Edge case handling**: You're working with unusual network structures that might challenge optimizations

### When to Use moneca_fast()

Choose the optimized `moneca_fast()` function when:

1. **Performance is important**: You're working with larger networks or need faster results
2. **Exploratory analysis**: You're doing initial exploration and can accept minor differences in segmentation
3. **Production environments**: You need reliable performance characteristics for automated analyses  
4. **Resource constraints**: You have limited computational resources or memory
5. **Large-scale studies**: You're processing multiple datasets and need efficiency

### Hybrid Approach

For comprehensive analysis, consider using both:

```{r hybrid-approach}
# Start with fast version for exploration
initial_result <- moneca_fast(small_data, segment.levels = 3, progress = FALSE)

# If results look promising and accuracy is needed, validate with original
if (length(initial_result$segment.list) > 2) {
  final_result <- moneca(small_data, segment.levels = 3)
  cat("Hybrid approach: Fast exploration followed by accurate validation\n")
}
```

## Quality Assessment

Both algorithms should produce similar high-level segmentation patterns, but may differ in details:

### Comparing Algorithm Outputs

```{r quality-assessment}
# Function to compare segmentation quality
compare_segmentations <- function(seg1, seg2, level = 2) {
  if (length(seg1$segment.list) < level || length(seg2$segment.list) < level) {
    return("Insufficient levels for comparison")
  }
  
  # Compare number of segments
  n_segments_1 <- length(seg1$segment.list[[level]])
  n_segments_2 <- length(seg2$segment.list[[level]])
  
  # Calculate segment size distributions
  sizes_1 <- sapply(seg1$segment.list[[level]], length)
  sizes_2 <- sapply(seg2$segment.list[[level]], length)
  
  list(
    n_segments = c(algorithm1 = n_segments_1, algorithm2 = n_segments_2),
    avg_segment_size = c(algorithm1 = mean(sizes_1), algorithm2 = mean(sizes_2)),
    size_variance = c(algorithm1 = var(sizes_1), algorithm2 = var(sizes_2))
  )
}

# Compare our previous results
comparison <- compare_segmentations(result_original, result_fast, level = 2)
print(comparison)
```

### Validation Strategies

1. **Cross-validation**: Run both algorithms and compare results
2. **Sensitivity analysis**: Test with different parameters
3. **Visualization**: Plot results to assess segmentation quality
4. **Domain expertise**: Validate results against substantive knowledge

```{r validation-example}
# Visualize results from both algorithms
if (requireNamespace("ggraph", quietly = TRUE)) {
  # Note: In practice, you would create actual plots here
  cat("Both algorithms can be visualized using plot_moneca_ggraph()\n")
  cat("Compare visual output for validation\n")
}
```

## Conclusion

Both `moneca()` and `moneca_fast()` implement the core MONECA methodology effectively, but with different optimization strategies:

### Summary of Key Differences

| Aspect | moneca() | moneca_fast() |
|--------|----------|---------------|
| **Clique Strategy** | All cliques | Maximal cliques |
| **Performance** | Slower, thorough | Faster, optimized |
| **Memory Usage** | Higher | Lower |
| **Accuracy** | Maximally conservative | Practically accurate |
| **Parameter Complexity** | Simple | More options |
| **Best Use Case** | Small networks, maximum accuracy | Larger networks, good performance |

### Recommendations

1. **Start with moneca_fast()** for initial exploration and most practical applications
2. **Use moneca()** when accuracy is critical or networks are small
3. **Compare results** when in doubt about which algorithm to trust
4. **Consider dataset characteristics** - sparse vs. dense, small vs. large
5. **Validate results** using domain knowledge and visualization

The choice between algorithms should be based on your specific requirements for accuracy, performance, and the characteristics of your mobility data. Both algorithms are valid implementations of the MONECA methodology and will provide valuable insights into mobility patterns and social structure.

## References

- Toubøl, J., & Larsen, A. G. (2017). Mapping the Social Class Structure: From Occupational Mobility to Social Class Categories Using Network Analysis. *Sociology*, 51(6), 1257-1276.

## Session Information

```{r session-info}
sessionInfo()
```