---
title: "Log-Linear Models for Mobility Tables"
author: "Giampaolo Montaletti"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Log-Linear Models for Mobility Tables}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)

# Check if required packages are available
required_packages <- c("ggplot2", "igraph", "stats")
missing_packages <- required_packages[!sapply(required_packages, requireNamespace, quietly = TRUE)]

if (length(missing_packages) > 0) {
  knitr::opts_chunk$set(eval = FALSE)
  message("Some required packages are missing: ", paste(missing_packages, collapse = ", "))
  message("Code examples will not be evaluated.")
}
```

## Introduction

A mobility table records how individuals move between categories (occupations, social classes, regions) over time. MONECA identifies clusters of categories with high internal mobility using a graph-theoretic algorithm. Log-linear models complement this by providing a statistical framework to test whether those clusters --- or other structural features --- capture genuine patterns in the data.

The moneca package provides two functions for this purpose:

- `fit_mobility_model()` fits a single log-linear model to a mobility table
- `compare_mobility_models()` compares multiple models side by side, with likelihood-ratio tests for nested pairs

This vignette walks through a complete analysis using synthetic data, covering the five model types, how to read their output, how to compare them, and how to inspect residuals for further insight.

## Setup: generate data and run MONECA

```{r data-setup}
library(moneca)

# Generate an 8-class mobility table
mx <- generate_mobility_data(
  n_classes = 8,
  n_total = 10000,
  immobility_strength = 0.6,
  class_clustering = 0.3,
  seed = 42
)

# Run MONECA
seg <- moneca_fast(mx, segment.levels = 3)
print(seg)
```

The resulting object `seg` contains segment assignments at multiple hierarchical levels. The log-linear models below test whether these assignments explain statistically meaningful structure in the table.

## The five model types

Each model represents a different hypothesis about what drives mobility patterns. They are ordered from simplest to most complex.

### Independence

The independence model assumes that origin and destination are unrelated: knowing where someone starts tells you nothing about where they end up. All cell frequencies are predicted from the row and column marginal totals alone.

```{r fit-independence}
m_ind <- fit_mobility_model(mx, type = "independence")
print(m_ind)
```

This model almost always fits poorly for mobility tables, because people tend to stay in their current position or move to nearby ones. It serves as a baseline against which other models are compared.

### Quasi-independence

The quasi-independence model adds a separate parameter for each diagonal cell, accounting for the fact that people tend to stay put (immobility). Off-diagonal flows are still assumed to follow the marginal distributions.

```{r fit-quasi-independence}
m_qi <- fit_mobility_model(mx, type = "quasi_independence")
print(m_qi)
```

The improvement over independence is typically large, because immobility is the dominant feature of most mobility tables. However, a low p-value or high dissimilarity index here means that the off-diagonal structure --- who moves where --- is not random.

### Quasi-symmetry

The quasi-symmetry model tests whether the flow from category A to B equals the flow from B to A, after accounting for differences in marginal sizes. It adds a parameter for each unordered pair of categories.

```{r fit-quasi-symmetry}
m_qs <- fit_mobility_model(mx, type = "quasi_symmetry")
print(m_qs)
```

If this model fits well, mobility is approximately symmetric: transitions between any two categories are balanced (relative to their sizes). If it fits poorly, some flows are directional --- people move more easily in one direction than the other.

The quasi-symmetry model is fitted using Iterative Proportional Fitting (IPF) rather than GLM, which makes it fast and numerically stable even for large matrices.

### Uniform association

The uniform association model adds a single parameter measuring the overall strength of ordinal association. It treats categories as ordered (by their position in the table) and asks: do adjacent categories exchange members more often than distant ones?

```{r fit-uniform-association}
m_ua <- fit_mobility_model(mx, type = "uniform_association")
print(m_ua)
```

This model is useful when categories have a natural ordering (e.g., social classes from lower to upper). If the mobility structure is multidimensional rather than ordinal, the single association parameter will not capture it well.

### MONECA topology

The MONECA topology model uses segment assignments from a MONECA analysis as structural predictors. It adds diagonal parameters (like quasi-independence) plus a segment-pair factor that distinguishes within-segment from between-segment flows.

```{r fit-topology}
m_top <- fit_mobility_model(seg, type = "moneca_topology", level = 2)
print(m_top)
```

If this model fits substantially better than quasi-independence, the segments identified by MONECA correspond to genuine mobility barriers: people move within segments more than expected and between segments less than expected.

## Reading the output

Each `fit_mobility_model()` call returns a `moneca_loglinear` object with several fields:

- **Deviance (G²)**: the likelihood-ratio statistic measuring how far the model is from the saturated model (which fits the data exactly). Lower values indicate better fit.
- **Degrees of freedom (df)**: the number of free data points minus the number of model parameters. More complex models use more parameters and have fewer degrees of freedom.
- **P-value**: derived from a chi-squared test of the deviance. A large p-value (> 0.05) means the model fits adequately; a small p-value means the model misses structure present in the data.
- **Dissimilarity index**: the proportion of cases that would need to be reclassified for the model to reproduce the observed table exactly. Values below 0.05 indicate good fit; below 0.03, excellent fit.
- **AIC / BIC**: information criteria for comparing non-nested models. Lower values are preferred. BIC penalizes complexity more heavily than AIC.
- **Fitted values** (`$fitted`): the model's predicted cell counts, as an n x n matrix.
- **Residuals** (`$residuals`): Pearson residuals, as an n x n matrix. Large positive residuals indicate flows the model under-predicts.

## Comparing models

The `compare_mobility_models()` function produces a summary table and likelihood-ratio tests in a single call. When passed a moneca object via `mx`, it automatically fits all five model types:

```{r compare-models}
comp <- compare_mobility_models(mx = seg, level = 2)
print(comp)
```

### Reading the comparison table

The table is sorted by BIC (lowest first). Key columns to examine:

- **BIC**: the primary criterion for model selection among non-nested models. The model with the lowest BIC provides the best trade-off between fit and parsimony.
- **Dissimilarity**: a more intuitive measure of practical fit. Compare values across models to see which ones capture the data most accurately.
- **P-value**: if any model has a large p-value, it fits the data adequately on its own. If all p-values are small, all models miss some structure, and the comparison focuses on which misses least.

### Reading the likelihood-ratio tests

The `lr_tests` component tests whether moving from a simpler model to a more complex one provides a statistically significant improvement. Each row shows:

- **model_1** (simpler) and **model_2** (more complex): the two nested models
- **delta_deviance**: the reduction in G² when moving from the simpler to the more complex model
- **delta_df**: the number of additional parameters used
- **p_value**: if small, the more complex model captures real structure that the simpler one misses

For example, a significant test between quasi-independence and MONECA topology means that the MONECA segments explain mobility patterns beyond what diagonal immobility alone can account for. The segments correspond to genuine structural features of the table, not artifacts of the clustering algorithm.

## Working with larger matrices

The log-linear functions scale to large matrices. The quasi-symmetry model uses IPF rather than GLM, so it remains fast even when GLM-based fitting becomes numerically expensive.

```{r large-matrix}
mx_large <- generate_mobility_data(
  n_classes = 30,
  n_total = 50000,
  immobility_strength = 0.6,
  class_clustering = 0.3,
  seed = 99
)

seg_large <- moneca_fast(mx_large, segment.levels = 3)

# Fit all models and compare
comp_large <- compare_mobility_models(mx = seg_large, level = 2)
print(comp_large)
```

For matrices with 50 or more categories, the quasi-symmetry model fitted via IPF typically converges in 20--50 iterations, while a GLM-based approach for the same model would require thousands of symmetric-pair parameters.

## Inspecting residuals

The residuals from the best-fitting model identify specific flows that the model does not capture well. Large positive Pearson residuals point to transitions that occur more often than the model predicts --- unexpected mobility channels that may warrant further investigation.

```{r residuals}
# Extract residuals from the MONECA topology model
resid_mx <- m_top$residuals

# Convert to a sorted data frame
resid_df <- data.frame(
  Origin = rep(rownames(resid_mx), ncol(resid_mx)),
  Destination = rep(colnames(resid_mx), each = nrow(resid_mx)),
  Residual = as.vector(resid_mx),
  stringsAsFactors = FALSE
)

# Remove diagonal (immobility is accounted for by the model)
resid_df <- resid_df[resid_df$Origin != resid_df$Destination, ]

# Show the largest under-predicted flows
resid_df <- resid_df[order(-resid_df$Residual), ]
head(resid_df, 10)
```

These are category pairs where observed mobility exceeds the model's prediction. They represent flows that cross the segment boundaries identified by MONECA --- transitions between clusters that the topology model does not anticipate. Such cross-segment flows can reveal labor market connections, institutional pathways, or structural changes that a cluster-based model misses.

## Summary

The typical workflow for log-linear analysis of a mobility table is:

1. **Run MONECA** (`moneca()` or `moneca_fast()`) to identify segment structure
2. **Compare models** with `compare_mobility_models()` to assess whether the segments capture genuine patterns
3. **Inspect residuals** from the best model to identify unexpected mobility channels

The five model types test different hypotheses:

| Model | What it tests | Typical use |
|:------|:-------------|:-----------|
| Independence | No association between origin and destination | Baseline |
| Quasi-independence | Immobility only; off-diagonal flows are random | Standard mobility baseline |
| Quasi-symmetry | Flows are balanced in both directions | Test for directional asymmetry |
| Uniform association | Single ordinal gradient drives mobility | Test for hierarchical ordering |
| MONECA topology | Segment structure explains mobility barriers | Validate MONECA segmentation |

When the MONECA topology model outperforms quasi-independence, it confirms that the algorithmic segmentation identifies real structural boundaries in the mobility data. The segments are not just a convenient visualization --- they correspond to statistically measurable discontinuities in how people move between categories.
