---
title: "Understanding MONECA Statistics: A Comprehensive Guide"
author: "MONECA Development Team"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Understanding MONECA Statistics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
library(MONECA)
```

## Introduction

The `print.moneca()` function provides a comprehensive summary of mobility analysis results from the MONECA clustering algorithm. This vignette explains each statistic in detail with clear definitions, interpretations, and practical examples.

## Overview of MONECA Output

When you run a MONECA analysis and print the results, you receive a structured report containing multiple sections of statistics. Let's start with a complete example:

```{r example-analysis}
# Generate synthetic mobility data
mobility_data <- generate_mobility_data(
  n_classes = 5,
  immobility_strength = 0.7,
  class_clustering = 0.8,
  seed = 123
)

# Run MONECA analysis
seg <- moneca(mobility_data, segment.levels = 3, small.cell.reduction = 5)

# Display results
print(seg)
```

## Statistics Explained

### 1. Overall Mobility Patterns

#### Overall Population Mobility Rate

**Definition:** The percentage of the total population that experiences any form of social mobility (movement between social classes).

**Calculation:** 
$$\text{Mobility Rate} = \frac{\text{Total Mobile Population}}{\text{Total Population}} \times 100$$

**Interpretation:**
- Values range from 0% (complete immobility) to 100% (everyone moves)
- Higher values indicate more fluid social structures
- Lower values suggest more rigid class boundaries
- Typical values in real data: 15-40%

```{r mobility-rate-example}
# Access the raw mobility rate
mx <- seg$mat.list[[1]]
l <- ncol(mx)
total_population <- mx[l, l]
mobile_population <- sum(mx[-l, -l])
mobility_rate <- mobile_population / total_population
cat("Raw mobility rate:", round(mobility_rate * 100, 1), "%\n")
```

#### Average Mobility Concentration

**Definition:** The mean percentage of all mobility that occurs through statistically significant pathways (edges with relative risk > 1) across all hierarchical levels.

**Interpretation:**
- Higher values (>80%) indicate mobility follows predictable, structured patterns
- Lower values suggest more random or dispersed mobility
- This metric helps assess whether mobility has clear "highways" or is more diffuse

### 2. Hierarchical Segmentation Analysis

#### Internal Mobility Within Segments

**Definition:** For each hierarchical level, the percentage of mobility that occurs within segments (clusters) rather than between them.

**Level Interpretation:**
- **Level 1:** Mobility within original social classes (baseline)
- **Level 2+:** Mobility within identified clusters/segments

**Pattern Analysis:**
- Increasing percentages across levels indicate successful clustering
- High values (>70%) at higher levels suggest strong segment cohesion
- If values don't increase, segments may not capture meaningful structure

```{r internal-mobility-detail}
# Show how internal mobility is calculated for Level 1
mx <- seg$mat.list[[1]]
l <- ncol(mx)
diagonal_sum <- sum(diag(mx)[-l])
total_mobility <- sum(mx[-l, -l])
internal_mobility_pct <- (diagonal_sum / total_mobility) * 100
cat("Level 1 internal mobility (manual calc):", round(internal_mobility_pct, 1), "%\n")
```

#### Mobility Concentration in Significant Pathways by Level

**Definition:** For each level, the percentage of mobility that flows through statistically significant pathways (those with relative risk > 1).

**Interpretation:**
- Shows how "channeled" mobility is at each level
- Higher values indicate more structured mobility patterns
- Useful for comparing how well each level captures mobility structure

### 3. Network Structure Metrics

```{r network-viz, fig.cap="Network visualization showing segments and connections"}
# Visualize the network at Level 2
plot_moneca_ggraph(seg, level = 2, node_color = "segment", 
                   edge_width_range = c(0.5, 3), layout = "fr")
```

#### Active Segments/Classes per Level

**Definition:** The number of distinct segments or classes at each hierarchical level.

**Pattern:**
- Decreases as segments merge at higher levels
- Rate of decrease indicates clustering aggressiveness
- Final count shows the fundamental mobility groups

#### Significant Edges After Cell Size Filtering

**Definition:** The count of statistically significant mobility pathways after removing those based on small cell sizes.

**Interpretation:**
- More edges = more complex mobility patterns
- Decrease across levels shows simplification of structure
- Zero edges would indicate completely isolated segments

#### Network Connectivity Density

**Definition:** The proportion of possible edges that actually exist in the network.

**Calculation:**
$$\text{Density} = \frac{\text{Actual Edges}}{\text{Possible Edges}}$$

**Interpretation:**
- Values range from 0 (no connections) to 1 (fully connected)
- Higher density indicates more interconnected mobility
- Often increases at higher levels as structure consolidates

#### Isolated Segments

**Definition:** The count of segments/classes with no statistically significant mobility connections to other segments.

**Interpretation:**
- Higher numbers indicate fragmented social structure
- Zero isolates means all segments are connected
- Isolates may represent truly closed social groups

### 4. Detailed Degree Distributions

When `show.degree.stats = TRUE` (default), additional network statistics are displayed:

```{r degree-stats-example}
# Show results without degree statistics for comparison
print(seg, show.degree.stats = FALSE)
```

#### Degree Types Explained

1. **Total Connections (All Degrees):** Sum of inward and outward connections
2. **Outward Mobility:** Number of destinations a segment sends people to
3. **Inward Mobility:** Number of origins a segment receives people from
4. **Edge Weights:** Strength of connections (relative risk values)

#### Reading the Distribution Tables

Each table shows six statistics:
- **Min:** Lowest value in the distribution
- **Q1:** First quartile (25th percentile)
- **Median:** Middle value (50th percentile)
- **Mean:** Average value
- **Q3:** Third quartile (75th percentile)
- **Max:** Highest value in the distribution

## Practical Interpretation Guide

### Step-by-Step Analysis

1. **Start with Total Mobility**
   - Is the system fluid or rigid?
   - Compare to known benchmarks or expectations

2. **Check Edge Concentration**
   - Are there clear mobility pathways?
   - Or is movement more random?

3. **Examine Diagonal Mobility Progression**
   - Does clustering reveal meaningful structure?
   - Are higher levels more cohesive?

4. **Review Network Metrics**
   - How connected is the system?
   - Are there isolated groups?

5. **Analyze Degree Distributions (if shown)**
   - Which segments are mobility hubs?
   - Which are more peripheral?

### Example Interpretation

Let's interpret our example results:

```{r interpretation-example}
# Create a different mobility pattern for comparison
mobility_data2 <- generate_mobility_data(
  n_classes = 5,
  immobility_strength = 0.9,  # Much less mobility
  class_clustering = 0.5,      # Less structured
  seed = 456
)

seg2 <- moneca(mobility_data2, segment.levels = 3, small.cell.reduction = 5)

cat("Comparison of two mobility systems:\n\n")
cat("System 1 (More fluid):\n")
print(seg, show.degree.stats = FALSE, digits = 1)

cat("\n\nSystem 2 (More rigid):\n")
print(seg2, show.degree.stats = FALSE, digits = 1)
```

## Advanced Usage

### Customizing Output

```{r custom-output}
# Show more decimal places for precise analysis
print(seg, digits = 2, show.degree.stats = FALSE)

# Adjust small cell reduction threshold
print(seg, small.cell.reduction = 10, show.degree.stats = FALSE)
```

### Accessing Raw Statistics

While `print.moneca()` provides formatted output, you can access the raw statistics for further analysis:

```{r raw-stats}
# The MONECA object contains all the raw data
str(seg, max.level = 2)

# Access specific components
cat("Number of levels:", length(seg$segment.list), "\n")
cat("Segments at level 2:", unique(seg$segment.list[[2]]), "\n")
```

## Common Patterns and Their Meanings

### Pattern 1: Strong Hierarchical Structure
- Internal mobility increases substantially across levels
- Network density increases at higher levels
- Few or no isolated segments
- **Interpretation:** Clear nested mobility structure

### Pattern 2: Weak Structure
- Internal mobility shows little increase
- Many isolated segments
- Low edge concentration
- **Interpretation:** Mobility is more random than structured

### Pattern 3: Polarized System
- High internal mobility even at Level 1
- Very low overall mobility rate
- High number of isolated segments
- **Interpretation:** Rigid class boundaries with little inter-class movement

## Troubleshooting

### Issue: Very High Network Density
**Cause:** Small cell reduction threshold too low
**Solution:** Increase `small.cell.reduction` parameter

### Issue: Many Isolated Segments
**Cause:** Threshold too high or genuinely fragmented structure
**Solution:** Check threshold or examine data for truly isolated groups

### Issue: No Change Across Levels
**Cause:** Data may lack hierarchical structure
**Solution:** Consider fewer levels or different clustering parameters

## Conclusion

The enhanced `print.moneca()` output provides a comprehensive view of mobility patterns and their hierarchical structure. By understanding each statistic and their relationships, researchers can:

1. Assess overall social fluidity
2. Identify structured mobility pathways
3. Discover hierarchical clustering in mobility patterns
4. Compare different populations or time periods
5. Validate theoretical models of social mobility

For more information on MONECA methodology, see `vignette("moneca-introduction")`. For visualization options, see the documentation for `plot_moneca_ggraph()`.