---
title: "Introduction to MONECA: Mobility Network Clustering Analysis"
author: "Jonas Touboel and Anton Grau Larsen"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_width: 8
    fig_height: 6
vignette: >
  %\VignetteIndexEntry{Introduction to MONECA: Mobility Network Clustering Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "figures/",
  warning = FALSE,
  message = FALSE
)

# Check if required packages are available
required_packages <- c("ggplot2", "ggraph", "igraph", "dplyr", "tidygraph")
missing_packages <- required_packages[!sapply(required_packages, requireNamespace, quietly = TRUE)]

if (length(missing_packages) > 0) {
  knitr::opts_chunk$set(eval = FALSE)
  message("Some required packages are missing: ", paste(missing_packages, collapse = ", "))
  message("Code examples will not be evaluated.")
}
```

## Introduction

moneca (Mobility Network Clustering Analysis) is an R package that analyzes social mobility patterns through graph-theoretic approaches. The package creates weighted networks from mobility tables and uses cliques to identify discrete and nested clusters of positions with high internal mobility.

### Key Features

- **Network-based clustering** of mobility tables
- **Hierarchical segmentation** at multiple levels
- **Modern visualization** with ggplot2 and ggraph
- **Quality metrics** for assessing segmentation performance
- **Synthetic data generation** for testing and demonstrations

## Basic Concepts

Social mobility can be represented as a network where:

- **Nodes** represent social positions (occupations, classes, etc.)
- **Edges** represent mobility flows between positions
- **Edge weights** represent the strength of mobility connections
- **Clusters** represent groups of positions with high internal mobility

MONECA identifies these clusters by:

1. Converting mobility tables to relative risk matrices
2. Finding cliques in the weighted network
3. Iteratively aggregating positions into segments
4. Creating hierarchical segmentation levels

## Installation and Setup

```{r installation, eval=FALSE}
# Install from GitHub (when available)
# devtools::install_github("antongrau/MONECA")

# Load the package
library(moneca)

# Load additional packages for visualization
library(ggplot2)
library(ggraph)
```

```{r load_package, echo=FALSE}
# For vignette building
library(moneca)
```

## Working with Synthetic Data

MONECA includes functions to generate synthetic mobility data for demonstrations and testing.

### Creating Basic Synthetic Data

```{r synthetic_basic}
# Generate a simple 5-class mobility table
basic_data <- generate_mobility_data(
  n_classes = 5,
  n_total = 2000,
  immobility_strength = 0.7,
  class_clustering = 0.2,
  class_names = c("Upper", "Upper-Mid", "Middle", "Lower-Mid", "Working"),
  seed = 42
)

# View the structure
print(basic_data)
```

The mobility matrix shows:
- **Rows**: Origin classes (where people start)
- **Columns**: Destination classes (where people end up)
- **Last row/column**: Marginal totals
- **Diagonal**: Immobility (staying in same class)

### Creating Different Mobility Scenarios

```{r synthetic_scenarios}
# Get multiple example datasets
examples <- generate_example_datasets()
names(examples)

# Examine the rigid class system
print("Rigid class system:")
print(examples$rigid)

# Examine the fluid (highly mobile) system
print("Fluid mobility system:")
print(examples$fluid)
```

## Running MONECA Analysis

### Basic Analysis

```{r basic_analysis}
# Run MONECA on the basic dataset
seg <- moneca(basic_data, segment.levels = 3)

# View the results
print(seg)
```

The MONECA object contains:
- `segment.list`: Hierarchical clusters for each level
- `mat.list`: Aggregated mobility matrices for each level

### Understanding the Results

```{r understand_results}
# View segment membership
membership <- segment.membership(seg)
print(membership)

# View the aggregated matrix for level 2
print("Level 2 aggregated matrix:")
print(seg$mat.list[[2]])
```

## Modern Visualization with ggraph

MONECA now includes modern visualization functions using ggplot2 and ggraph.

### Basic Network Plot

```{r plot_basic, eval=TRUE}
# Create a basic network plot
p1 <- plot_moneca_ggraph(
  seg,
  title = "Basic MONECA Network",
  node_color = "segment",
  node_size = "total"
)

print(p1)
```

### Customized Visualization

```{r plot_custom, eval=TRUE}
# Create a customized plot with different layout and styling
p2 <- plot_moneca_ggraph(
  seg,
  layout = "kk",  # Kamada-Kawai layout
  node_color = "segment",
  edge_width = "weight",
  color_palette = "Spectral",
  theme_style = "minimal",
  title = "Customized MONECA Visualization",
  show_labels = TRUE
)

print(p2)
```

### Ego Network Analysis

Ego networks show mobility patterns from a single position:

```{r ego_plot, eval=TRUE}
# Create ego network for the middle class (position 3)
p3 <- plot_ego_ggraph(
  seg,
  basic_data,
  ego_id = 3,
  title = "Mobility from Middle Class",
  highlight_color = "red"
)

print(p3)
```

### Stair Plots: Visualizing Segmentation Levels

```{r stair_plot, eval=TRUE}
# Create stair plots showing different segmentation levels
stair_plots <- plot_stair_ggraph(seg, levels = 2:3)

# Display individual plots
print(stair_plots$`Level 2`)
print(stair_plots$`Level 3`)
```

## Advanced Analysis Examples

### Comparing Different Mobility Regimes

```{r compare_regimes}
# Analyze the rigid class system
rigid_seg <- moneca(examples$rigid, segment.levels = 3)

# Analyze the fluid mobility system  
fluid_seg <- moneca(examples$fluid, segment.levels = 3)

# Compare segment memberships
print("Rigid system segments:")
print(segment.membership(rigid_seg))

print("Fluid system segments:")
print(segment.membership(fluid_seg))
```

### Visualizing Different Mobility Regimes

```{r compare_plots, eval=TRUE}
# Plot rigid system
p_rigid <- plot_moneca_ggraph(
  rigid_seg,
  title = "Rigid Class System",
  node_color = "segment",
  color_palette = "Reds"
)

# Plot fluid system
p_fluid <- plot_moneca_ggraph(
  fluid_seg,
  title = "Fluid Mobility System", 
  node_color = "segment",
  color_palette = "Blues"
)

print(p_rigid)
print(p_fluid)
```

## Analytical Functions

### Segment Quality Assessment

```{r quality_assessment}
# Assess the quality of segmentation
# (Note: this function may need to be implemented)
# quality_metrics <- segment.quality(seg)
# print(quality_metrics)

# Alternative: examine within vs. between segment mobility
level2_matrix <- seg$mat.list[[2]]
print("Level 2 mobility matrix (aggregated):")
print(level2_matrix)

# Calculate internal mobility rates
internal_mobility <- diag(level2_matrix[-nrow(level2_matrix), -ncol(level2_matrix)])
total_mobility <- level2_matrix[-nrow(level2_matrix), ncol(level2_matrix)]
immobility_rates <- internal_mobility / total_mobility

print("Immobility rates by segment:")
print(immobility_rates)
```

### Edge Analysis

```{r edge_analysis}
# Examine edge weights between segments
edges <- segment.edges(seg, level = 1)
print("Edge weights matrix:")
print(edges)

# Focus on strongest connections
strong_edges <- edges
strong_edges[strong_edges < 2] <- NA
print("Strong connections (relative risk > 2):")
print(strong_edges)
```

## Working with Real Data

When working with your own mobility data, ensure it follows the MONECA format:

### Data Format Requirements

```{r data_format}
# Your data should be a matrix with:
# - Rows: origin positions
# - Columns: destination positions  
# - Last row: column totals
# - Last column: row totals
# - Bottom-right cell: grand total

# Example of properly formatted data structure:
example_structure <- matrix(c(
  100, 20, 10, 130,  # Origin 1: 100 stay, 20 to pos 2, 10 to pos 3, total 130
  15, 200, 25, 240,  # Origin 2: 15 to pos 1, 200 stay, 25 to pos 3, total 240  
  10, 30, 180, 220,  # Origin 3: 10 to pos 1, 30 to pos 2, 180 stay, total 220
  125, 250, 215, 590 # Totals: 125 end in pos 1, 250 in pos 2, 215 in pos 3, grand total 590
), nrow = 4, ncol = 4)

rownames(example_structure) <- c("Class_A", "Class_B", "Class_C", "Total")
colnames(example_structure) <- c("Class_A", "Class_B", "Class_C", "Total")

print("Properly formatted mobility matrix:")
print(example_structure)
```

### Data Validation

```{r data_validation}
# Check if your data is properly formatted
validate_mobility_matrix <- function(mat) {
  n <- nrow(mat) - 1
  
  # Check row totals
  calculated_row_totals <- rowSums(mat[1:n, 1:n])
  actual_row_totals <- mat[1:n, n + 1]
  
  # Check column totals  
  calculated_col_totals <- colSums(mat[1:n, 1:n])
  actual_col_totals <- mat[n + 1, 1:n]
  
  # Check grand total
  calculated_grand_total <- sum(mat[1:n, 1:n])
  actual_grand_total <- mat[n + 1, n + 1]
  
  list(
    row_totals_match = all.equal(calculated_row_totals, actual_row_totals),
    col_totals_match = all.equal(calculated_col_totals, actual_col_totals),
    grand_total_matches = all.equal(calculated_grand_total, actual_grand_total)
  )
}

# Validate our example
validation <- validate_mobility_matrix(example_structure)
print(validation)
```

## Best Practices and Tips

### Choosing Parameters

1. **Number of levels**: Start with 3-4 levels, adjust based on your data size
2. **Cut-off values**: Default (1.0) works well for relative risk; adjust for sensitivity
3. **Small cell reduction**: Use to handle sparse data in large matrices

### Interpretation Guidelines

1. **Segments represent groups** with high internal mobility
2. **Hierarchical structure** shows nested clustering at different scales  
3. **Edge weights** indicate relative risk compared to independence
4. **Layout algorithms** can emphasize different network properties

### Performance Considerations

1. **Large matrices** (>50x50) may require increased computation time
2. **Progress bars** show optimization progress for large datasets
3. **Memory usage** grows with number of levels and matrix size

## Conclusion

MONECA provides a powerful framework for analyzing social mobility through network-based clustering. The combination of:

- Flexible clustering algorithms
- Modern visualization capabilities  
- Hierarchical segmentation
- Quality assessment tools

Makes it suitable for exploring mobility patterns in diverse social contexts.

### Further Reading

- Original MONECA methodology papers
- Network analysis in social mobility research
- Graph-theoretic approaches to social structure

### Getting Help

- Use `?function_name` for detailed function documentation
- Check the package website for updates and examples
- Report issues on the GitHub repository

```{r session_info}
# Session information for reproducibility
sessionInfo()
```